---
title: "03-lab03"
output:
  html_document:
    css: webex.css
    includes:
      after_body: webex.js
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(webex)
```

```{block, type = "info"}

In lab 2, you were introduced to the R environment (e.g. setting your working directory and the difference between .R and .Rmd files). You also began working with messy data by having a go at loading in datasets using `read_csv()`, joined files together using `inner_join()`, and pulled out variables of interest using `select()`. 

In lab 3, we'll be moving on to becoming familiar with the Wickham Six and the functionality of the R package, `tidyverse`! 

```

## Pre-Class Activity 

**Prep Materials**

* Watch Heather's "Intro to `tidyverse`" video 
* Check out the [Level One Psychology Practical Support]() website (designed by our very own GTAs!) for more information on `tidyverse` and its powerful functions 
*  Work through the tidyverse book which talks you through the purpose of tidyverse using examples on a dataset called `babynames`. This package is installed on the lab computers but you can also `install.packages("babynames")` on your own personal computer if you want to prep for the lab at home

**Pause here and test your knowledge from lab 2** 

* How do you load in `tidyverse()`? 

`r mcq(c(answer = "library(tidyverse)","install.packages(tidyverse)"))`
<br>

* What would be the correct code to successfully combine `dat` and `pinfo`? 

`r mcq(c(answer = "inner_join(dat, pinfo, by = 'id', 'intervention')", "select(dat, pinfo, by = 'id', 'intervention'", "inner_join(dat, pinfo, by = 'id')"))`


## What is `tidyverse`? 

`tidyverse` is a collection of R packages with a common design philosophy, grammar, and data structure that makes analysis faster, easier, and (if you can believe it) fun!

`tidyverse` contains six core packages: dplyr , tidyr , readr ,  purrr ,  ggplot2 , and tibble. If you've ever typed library(tidyverse) into R, you will have seen that it loads in all of these packages in one go. Within these six core packages, you should be able to find everything you need to analyse your data. 

One of the key points about `tidyverse` is that all data structures follow a **tidy** format:

1. Each variable is in a **column** (e.g. name)
2. Each observation is a **row** (e.g. each individual name is on its on row)
3. Each value is a **cell** (e.g. each individual name is contained in its own cell)

If you make sure your data is tidy, you’ll spend less time rummaging through rubbish on your research path and spend more time working on your all important analysis!

## The Wickham Six 

Hadley Wickham, a driving force behind `tidyverse`, argues that 90% of data analysis can be reduced to six important functions:

* `select()`: include or exclude certain certains
* `filter()`: include or exclude certain rows 
* `mutate()`: create new columns and make changes to existing ones
* `arrange()`: change the order of rows (ascending or descending)
* `group_by()`: organise observations into groups 
* `summarise()`: provide summary statistics for groups of observations (e.g. mean, median) 

These six functions are known as ’single table verbs’ because they only operate on one table at a time. Later in the course you will learn two-table verbs that you can use to merge tables together. Although the operations of these functions may seem very simplistic, it’s amazing what you can accomplish when you string them together: Hadley Wickham has claimed that 90% of data analysis can be reduced to the operations described by these six functions. 

******
**Remember these key points as you work through the data activities on this course**

* RStudio is a tool that enables you to become **confident** and **competent** working with data
* This is **essential** as a psychologist as you need to know that the data findings are based on are reliable and valid. Imagine delivering a treatment that was based on research where the analyst was not confident or competent?
* **Think of a skill you nailed the first time you did it**. Not so easy? You will learn to work with data by repeating these basic skills over and over. It's new and will take time. Give yourself credit for trying and don't stress if it doesn't work right away. Use the resources we give, that are being provided online among the open science community, and ask questions in class and on the forums. **Even the GTAs supporting you in the labs felt overwhelmed at first!** Listen to them talk about how their challenges learning R made their successes even better [here](https://psyteachr.github.io/hack-your-data/podcast.html)

******

## In Class Activity 

Open up a new script (**File > New File > R Script**).  What is our first step at the start of any new R session? 

`r hide("I need a hint")`

We set our **working directory**!

Take some time now to make sure you have your script file and data in the same folder, set your working directory to this folder. 

`r unhide()`

### Activity 1: Packages and Data

After we have set our working directory, it's time to load in `tidyverse` and our data. Think of R like a smartphone - you can't just pick up your phone and open Instagram without actually carrying out any actions to open up the app (i.e. clicking on the icon). In the same way, R won't start using `tidyverse` until you tell it to! 

```{r T1, warning = FALSE, message = FALSE}

library(tidyverse) 

dat <- read_csv('files/ahi-cesd.csv')
pinfo <- read_csv('files/participant-info.csv')
all_dat <- inner_join(dat, pinfo, by='id', 'intervention')

```

Now let's start working with our `tidyverse` verb functions...

******

Needing some clarification on `tidyverse` verb functions? Check out the [Level One Psychology Practical Support Page](https://shannon-mcnee.github.io/UG-level1-prac/)

Remember - you can also get help from R at any time by typing `?function name` into your console!

******

### Activity 2: Select  

Select the columns all_dat, ahiTotal, cesdTotal, sex, age, educ, income, occasion, elapsed.days from the data and create a variable called ```summarydata```. 

```{r, T2}
summarydata <- select(all_dat, ahiTotal, cesdTotal, sex, age, educ, income, occasion, elapsed.days)
```

******

**Pause here and interpret the above code and output**

* Discuss in pairs or groups how you would translate this code into English 
* What columns have been removed from the data? 

******

### Activity 3: Arrange  

Arrange the data in the variable created above (```summarydata```) by ahiTotal with lowest score first. 
```{r, T3}
ahi_asc <- arrange(summarydata, by = ahiTotal)
```

******

**Pause here and interpret the above code and output**

* How could I arrange this data in **descending** order (highest score first)?  

******

### Activity 4: Filter  

Filter the data ```ahi_desc``` by taking out those who are over 65 years of age.  
```{r, T4}
age_65max <- filter(ahi_asc, age < 65)
```

******

**Pause here and interpret the above code and output** 

* What does `filter()` do? 

`r mcq(c("splits a column into multiple columns", "transforms existing columns", "takes multiple columns and collapses them together", answer = "removes information that we are not interested in"))`

******

### Activity 5: Summarise  

Then, use summarise to create a new variable ```data_median```, which calculates the median ahiTotal score in this grouped data and assign it a table head called ```median_score```.
```{r T5}
data_median <- summarise(age_65max, median_score = median(ahiTotal))
```

******

**Pause here and interpret the above code and output**

* How could I change this code to give me the **mean score**?

******

### Activity 6: Group_by  

Group the data stored in the variable ```age_65max``` by sex, and store it in ```data_sex``` then use mutate to create a new column called Happiness which categorises participants based on whether they score above the median ahiTotal score for females.

```{r T6}
data_sex <- group_by(age_65max, sex)
happy_female <- mutate(data_sex, Happiness_Female = (ahiTotal > 74))
```

******

**Pause here and interpret the above code and output**

* What does `group_by()` do? 

`r mcq(c("provides summary statistics of an existing dataframe","organises information in ascending or descending order", "transforms existing columns", answer = "groups data frames based on a specific column so that all later operations are carried out on a group basis"))`

******




