# Lab 6

## Pre-class activity 
Students were directed to labs 4 and 5 to recap on skills needed before applying them to this data set introduced in lab 5. 

## Inclass activity 

### Task 1 - loading in the package 
We are going to be working with the Autism Quotient (AQ) data which you have already been introduced to in lab 1 and in the pre-class for this lab. The relevant data required to use this data comes in 3 seperate files; the survey responses 'responses.csv', the question formats 'qformats.csv' and scoring info 'scoring.csv'. To get us started, load in tidyverse.
```{r T1, message = FALSE}
# load in tidyverse by replacing this line
library(tidyverse)
```

### Task 2 - read in data
Read in the data to create ```responses```, ```qformats``` and ```scoring``` using the corresponding data files for each tibble i.e. responses for the response data, qformats for the question formats and scoring for the scoring info.
```{r T2, message = FALSE}
responses <- read_csv("files/responses.csv") 
qformats <- read_csv("files/qformats.csv") 
scoring <- read_csv("files/scoring.csv") 
```

**Checkpoint** Has the data now appeared in the environment? Take a look and make sure they look as you would expect them to. Do they look the same as they did in the .csv files? Is it in the format ideal for analysis (see lab 1 in-class and lab 2 prep for a clue)?

### Task 3 - getting our data into the correct format for analysis 
Let's get this data into long format from wide format.Firstly, create ```rlong``` by using gather to show the question and response for each participant in long format. 
```{r T3}
rlong <- gather(responses, Question, Response, Q1:Q10)
```

### Task 4 - Question format
Now use inner_join to add in the question format (QFormat) for each question and store it in ```rlong2```. This should now have 4 columns showing Id, Question, Response and QFormat.
```{r T4}
rlong2 <- inner_join(rlong, qformats, "Question")
```

### Task 5 - Scores
We can now get the score for each participant for each question. Use inner_join to add in the score for each question and store it in ```rscores```. 
```{r T5}
rscores <- inner_join(rlong2, scoring, c("QFormat", "Response"))
```

### Task 6 - Calculate AQ scores
To calculate the AQ score for each participant, firstly we need to group the scores by Id. Using the tibble `rscores` defined in task 5, group the AQ scores by Id and store it in aq_scores1.
```{r Q6}
aq_scores1 <- group_by(rscores, Id)
```

### Task 7 - Calculate AQ scores cont'd
Now we need we have the scores grouped by Id we can use summarise to calculate the sum of each question's score for each participant. Do this for ```aq_scores1``` and store it in ```aq_scores2```.  
```{r Q7}
aq_scores2 <- summarise(aq_scores1, AQ = sum(Score))
```

### Task 8 - Calculate AQ scores efficiently
There is a more efficient way of calculating the AQ scores for each participant by combining the lines of code in tasks 6, 7 and 8 together into one tibble using pipes. *Using the tibble `rscores` defined in task 5*, calculate AQ scores for each participant. Name this table aq_scores (USE THIS EXACT NAME!)
```{r Q8}
aq_scores <- rscores %>%
  group_by(Id) %>%
  summarise(AQ = sum(Score))
```

**Checkpoint** Does ```aq_scores``` look the same as ```aq-scores2```?

### Task 9 - Using ggplot2, make a distribution of the overall AQ scores by producing a histogram from aq_scores.
```{r Q9, message = FALSE}
## TODO: uncomment the following line, and replace it with your answer:
ggplot(aq_scores, aes(AQ)) + 
  geom_histogram()
```

#### Well done! You should use this activity to help you complete the assessment to be submitted 1 week from your lab by 12pm. 